#lab_11
Rob Wells, April 11, 2024

## About this lab

To complete this lab, you need to:
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Overview

There are three main goals of this lab:

* learn to make use of R packages that can access data via third-party APIs.
* learn to develop a quick and accurate understanding of what data can be obtained via third-party APIs. 
* learn to develop questions that can be answered using that data, and avoid developing questions that cannot. 

## APIs to choose from
 

* tidycensus - for asking census-related questions.

## Load libraries

**Task**: load the tidyverse and janitor.  Uncomment the one other package you plan to work with.  Do not choose more than one API package.

```{r}
library(tidyverse)
library(janitor)
library(tidycensus)

```

## Get setup to access third-party APIs

### tidycensus

Use the U.S. census access token you created for this week's lab.  You can store it by editing the block below to replace 'YOUR KEY HERE' with your actual key, a long string of letters and numbers. If you need to sign up for another one, you can do so here:  [apply for an API key with the Census Bureau](https://api.census.gov/data/key_signup.html). 

```{r}
census_api_key("4a48d6839c89b310e63a151db25f4787b1b3e5de", install=TRUE, overwrite = TRUE)
```
05ded8bea43a2ceafcf1942349622055ff983157


### tidycensus

This package is designed to help you pull demographic information about different U.S. geographies.

* Basic usage https://walker-data.com/tidycensus/articles/basic-usage.html
* Functions: https://walker-data.com/tidycensus/reference/index.html
* More details https://walker-data.com/tidycensus/

Sample question: which state saw the largest growth in Hispanic population between 2010 and 2020?

```{r}

###
# Load ACS variables
###
acs_vars <- load_variables("acs5", year=2020)

###
# Get median household income for Maryland 2020
###

md <- get_acs(geography = "county",
              variables = c(median_income = "B19013_001"),
              state = "MD",
              year = 2020) 


```

### Develop and answer questions

**Task** Develop two different *interesting* questions that you can answer with the data. Run them by your professor before getting started on answering, to ensure they're sufficiently interesting and possible with the data. 

1. What U.S. counties have the highest growth in young population (24-29 years old) in the past 10 years?
2. What is the percentage of the whole US population that is covered by health insurance? 

**Task** Write code to answer the questions.  Write up a sentence with each of your findings. Produce at least one visualization to support one of your findings. 

1. What U.S. counties have the highest growth in young population (24-29 years old) in the past 10 years?    
```{r}
youngpeople2010 <- get_acs(geography = "county",
              variables = c(twentyfive_twentynine ="B07001_006", twenty_twentyfour ="B07001_005"),
              year = 2010) %>% 
              select(NAME, variable, estimate) %>% 
              pivot_wider(names_from = variable, values_from = estimate) %>% 
              mutate(pop_2010 = twenty_twentyfour + twentyfive_twentynine) 

youngpeople2020 <- get_acs(geography = "county",
              variables = c(twentyfive_twentynine ="B07001_006", twenty_twentyfour ="B07001_005"),
              year = 2020) %>% 
              select(NAME, variable, estimate) %>% 
              pivot_wider(names_from = variable, values_from = estimate) %>% 
              mutate(pop_2020 = twenty_twentyfour + twentyfive_twentynine) 

youngpeople2010_2020 <- youngpeople2010 %>% 
  inner_join(youngpeople2020, by=c("NAME")) %>% 
  mutate(pct_change = (pop_2020 - pop_2010)/pop_2010 *100) %>% 
  arrange(desc(pct_change)) %>% 
  head(5) %>% 
  select(NAME, pop_2010, pop_2020, pct_change)

print(youngpeople2010_2020)
#The US counties with the biggest percent increase in young people from 2010-2020 were Denali Borough, Alaska at 470.3%, Hooker County, Nebraska at 264.3%, Sweet Grass County, Montana 218.9%, Yakutat City and Borough, Alaska at 212.5% and McKenzie County, North Dakota at 210.4%.

```


2. What is the percentage of the whole US population that is covered by public health insurance in 2020 and the top 5 covered counties? 
```{r}
insured_2020 <- get_acs(geography = "county",
              variables = c(covered = "B992703_002"),
              year = 2020) %>% 
              mutate(public_insured_2020 = estimate)


population_2020 <- get_acs(geography = "county",
              variables = c(covered = "B01003_001"),
              year = 2020) %>% 
              mutate(pop_2020 = estimate)

total_insured_2020 <- sum(insured_2020$estimate)
total_population_2020 <- sum(population_2020$estimate) 
 
total_insured_2020/total_population_2020 *100
#According to this, 15.32335% of people were covered by public health insurance in 2020

pct_insured <- insured_2020 %>% 
  inner_join(population_2020, by=c("NAME", "GEOID")) %>% 
  mutate(pct_covered = public_insured_2020/pop_2020 *100) %>% 
  arrange(desc(pct_covered)) %>% 
  head(5) %>% 
  select(NAME, public_insured_2020, pop_2020, pct_covered)

print(pct_insured)
#The counties with the highest percentage of public insurance coverage in 2020 were McLean County, North Dakota at 41.4%, Lynn County, Texas at 36.3%, Wells County, North Dakota at 35.9%, Graham County, Arizona at 35.3% and Greenlee County, Arizona at 33.2%.
```
