#lab_08
Sean Mussenden, Aug 16, 2022

## About this lab

To complete this lab, you need to:
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings
**Task**: Load janitor and the tidyverse

```{r}
options(scipen=999)
library(tidyverse)
library(janitor)
```
## Load data

```{r}
md_black <- read.csv("maryland_black.csv")
md_hisp <- read.csv("maryland_hispanic.csv")
md_white <- read.csv("maryland_white.csv")
md_pop <- read.csv("maryland_population.csv")

```



**Task**: Load four tables of U.S. Census data.  Each table has one row per Maryland county with 2020 population information.  Each table has these same two columns:

* geoid - a unique identifier for each county
* name - the county name

The third column in each table represents the total population in one of four demographic categories.

* maryland_population.csv - total_pop - total population 
* maryland_hispanic.csv - hispanic_pop - total Hispanic population
* maryland_white.csv - white_pop - total White alone population
* maryland_black.csv - black_pop - total Black alone population

Some of the information in the tables may need to be cleaned or validated in order to produce the correct answer.  

## Questions

**Q1**  What Maryland county has the largest Hispanic population (by percentage of total population)? 

```{r}

md_hisp <- md_hisp %>%
  mutate(name = case_when(name == "Montgomery County,Maryland" ~ "Montgomery County, Maryland", TRUE ~name))

md_master <- md_pop %>%
  inner_join(md_black) %>% 
  inner_join(md_white) %>% 
  inner_join(md_hisp)

md_master <- md_master %>% 
  mutate(pct_hisp = hispanic_pop/total_pop *100)
```
**A1** Montgomery County had the largest Hispanic population at 19.5%


**Q2**  What is the total population of people who identify as white alone in Maryland? 
```{r}

#Somerset's white population of 13510000 needs to be divided by 100

md_white <- md_white %>%
  mutate(white_pop = case_when(
    white_pop == "13510000" ~ 13510000/100,
    TRUE ~ white_pop
  ))

md_master <- md_pop %>%
  inner_join(md_black) %>% 
  inner_join(md_white) %>% 
  inner_join(md_hisp)

sum(md_white$white_pop)

```
**A2**  3396638 people identified as white in MD in 2020.

**Q3** What is the average Black population in all Maryland counties? 
```{r}
#Link https://www.census.gov/quickfacts/fact/table/baltimorecitymaryland/RHI225222#RHI225222 
#According to the Census Bureau, Baltimore City's April 1, 2020 population was 585,708 and 61.2% of the population was Black, so this is 358,453 people.
#According to the Census Bureau, Baltimore County's April 1, 2020 population was 854,535 and 31.9% of the population was Black, so this is 272,596 people.

md_black <- md_black %>% 
  mutate(black_pop = (case_when(
    name == "Baltimore County, Maryland" ~ 272596, 
    name == "Baltimore city, Maryland" ~ 358453,
    TRUE ~ black_pop
)))

mean(md_black$black_pop)
```
**A3** The average Black population in MD counties is 75669.88.
