#Lab_04 {lab_04}

rob wells, 2/13/2024

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse here
library(tidyverse)

# Load janitor here
library(janitor)


library(dplyr)
```

### Load data


In this data set, each row represents a city or town in Maryland with the 2010 and 2020 figures for: 

-  Total population 
-  White only population
-  Black or African American only population
-  Hispanic only population 
-  Native American only population
-  Asian only population
-  Pacific Islander (including native Hawiian) only population

Background on this data from [Social Explorer](https://www.socialexplorer.com/data/ACS2016_5yr/metadata/?ds=ACS16_5yr&table=B02001) 
```{r}
# Write code to load the Maryland race by city, place for 2010 - 2020 and call it md_race.
md_race<- read_csv("lab_04.csv")


# The data is called lab_03.csv, and is in the same folder as lab_03.qmd (this data notebook)

```

## Answer questions

Q1: Name the five cities that had the largest Hispanic population as a percentage of the entire population in 2020 and call it top5_hispanic. Produce a top 5 table for each, selecting the place, 2020 Hispanic, 2020 total and your new calculation.

A1:
```{r}

pct_x2020_hispanic <- md_race %>%
  select(place, x2020_hispanic, x2020_total) %>%
  mutate(pct_x2020_hispanic = (x2020_hispanic/x2020_total)*100) %>% 
  slice_max(pct_x2020_hispanic, n=5)
View(pct_x2020_hispanic)

The five MD citiies with the largest percentage of hispanic population in 2020 were Langley Park (82%), Templeville town (77%), Marydel town (71%), Henderson town (64%) and Edmonston town (64%).

```

Q2: Compare the top 10 cities with the largest Black population as a percentage of the overall population in 2010 and 2020. Filter for places with more than 20,000 population in 2010. Call your results pct_black. Compare results for 2020 vs 2010. What is different? What surprised you? 

A2:

```{r}
pct_black <- md_race %>%
  select(place, x2010_total, x2010_black, x2020_total, x2020_black) %>%
   filter(x2010_total > 20000) %>%
  mutate(pct_black_2010 = (x2010_black/x2010_total) *100,
         pct_black_2020 = (x2020_black/x2020_total) *100) %>%
  slice_max(pct_black_2020, n=10)


What changes wasn't consistent over the cities. Some had different drops and increases in the total Black population. I was surprised to see that Baltimore City's Black popultion decreased from 2010 to 2020, even though it was only by 2 percentage points.

```

Q3: Find the 15 communities that had the great gains in Hispanic population from 2010 to 2020 as measured by percentage change. Also calculate percentage change for the community's overall growth during the same period. Filter your results for communities with more than 10,000 people as of 2020. Your results will be hisp_growth. Look at your results, craft a one-paragraph story pitch for your editor. 

A3:

```{r}
hisp_growth <- md_race %>%
  select(place, x2010_hispanic, x2010_total, x2020_hispanic, x2020_total) %>%
   filter(x2020_total > 10000) %>%
  mutate(hisp_growth = 100*(x2020_hispanic - x2010_hispanic)/x2010_hispanic,
         total_growth = 100*(x2020_total - x2010_total)/x2010_total)%>%
  slice_max(hisp_growth, n=15)

This data could be misleading, so I want to look into one of the top 15 places for growth: Urbana. Accoridng to Frederick County Public Schools, Urbana High School (which is a majority white) is only 12.8% Hispanic/Latino. Still, this data shows that Urbana's hispanic growth was an 160% increase, which may lead a reader to think that Urbana has a large hispanic population. I want to reach out to Urbana HS and other local groups to see what changes have been made in the community facing this 160% increase and if the community actually considers itself to have a growing Hispanic or large Hispanic population or if that is a misleading narrative from the data. 
```
